Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences

### 2016-10-06 paper/paper.tex
first draft long CogSci version

\subsection{Discussion}
Broadly speaking these results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome selective pressures and stabilize in a population provided there is a bias for simpler representations. This prediction hinges on three assumptions. First, that language is pressured toward both expressivity and learnability. Second, that language use is not too unpredictable: low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn. Third, that learners prefer simpler over more complex lexical representations. An important remark on this third condition is that a combination of rationality in choice and maximization in learning requires a weaker bias towards simpler representations. Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.  This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer.

While a major share of non upper-bounded lexical meanings is predicted by the literature, it is less clear to what extent other types should be present in the final population -- if at all. There are strong reasons to expect functionally suboptimal types $L_1$-$L_3$ to be ruled out since they fail to enable their users to communicate effectively. However, this is not true of $L_4$.\footnote{$L_6$ presents a special case. In our current setup, it mirrors $L_5$ in enabling for a pragmatic strengthening of message $m_2$ rather than $m_1$. However, this association of $s_2$ with $m_2$ under favorable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$. $L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'. However, our representation of lexica as Boolean matrices is blind to this.} Notwithstanding, it is not unthinkable that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another (\hl{references on individual level differences in the derivation of scalar implicatures}). Alternatively, a stronger tendency for posterior maximization has to be assumed (cf. Figure \ref{fig:prior-posterior}). This empirical issue relates to other two aspects left undiscussed in our present analysis: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences. We tacitly assumed pragmatic reasoning to come at no cost. However, there is experimental evidence that suggests  that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf. \citealt{deNeys+schaeken:2007, huang+snedeker:2009}. This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations. Should cost play a role then its effect is bound to depend on the frequency with which a given scalar expression is used. We conject that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line. In turn, this opens a possible venue to address the previous question about expected use of $L_4$-like semantics.

\section{General discussion}
We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of learning biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}. This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on language. We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer instead may lie in learnability: Simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred. This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences fail to lexicalize.

The main innovations of the model are its modular separation of expressivity and learnability, which allows for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning. The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}. In contrast to the present proposal, however, expressivity is modelled as exerting its force in the production of learning data only. This model's expressivity parameter thereby fulfills a similar role  to high values of $\lambda$ and $\alpha$ in making speaker behavior more deterministic. Consequently, it ``favors'' unambiguous languages. Crucially, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration. Our reasoning behind its inclusion lies in the empirical and theoretical observation that learnability alone can have the undesirable outcome of favoring functionally defective languages. This outcome was showcased for a tautological language in our analysis but similar effects have been reported in a number laboratory experiments with human subjects where the participants' task was to learn and subsequently reproduce the language produced by a previous participant, leading to a proliferation of languages were a large number of meanings are associated with a single form (\hl{references}). In contrast, adding an interactive component to the experimental setup has been found to foster languages that enable interlocutors to distinguish meanings accurately  (\hl{references}; for a review of these studies and further references see \citealt{kirby+etal:2015, tamariz+kirby:2016}). More generally, there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer. The former solely depends on the mechanism by which speakers associate form and meaning. The latter additionally hinges on the hearers linguistic experience and her capability to interpret linguistic input based on this experience. We contend that successful information transfer in a linguistic community is central the adoption of a communication system and that this measure is not faithfully reflected by taking only production into consideration.

A central assumption to the model's predictions is the learning bias for simpler lexical representations. Albeit implicit in our formalization, the conceptual foundation for a compression preference at the lexical level is grounded in work in the language of thought tradition put forward by \citet{piantadosi+etal:2012a,piantadosi+etal:underreview}. Following Piantadosi and colleagues we assumed that lexical meanings are expressed in a representational language and that the learners' task is to infer lexica that best explain the linguistic witnessed. Furthermore, we assumed learners to not only infer lexical meaning but types of linguistic use. Whether language use is culturally transmitted is an empirical matter that can not be settled here. However, this idealized model suggests that semantic structure evinced cross-linguistically that lends itself for pragmatic exploitation reflects lexical information that is both easy to learn and, more interestingly, also easy to learn how to exploit via mutual reasoning.

\section{Conclusion}
Language change is affected by intertwined pressures. We investigated the selection of non-upper-bounded lexical meanings as a case study on role of semantics and pragmatics in satisfying such pressures. We showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use. That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages fail to lexicalize certain pragmatic inferences. More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interaction as well as their representational complexity.

@@

\section{Some thoughts + what may be missing}
\begin{itemize}
  \item I still need to review, read and find some references I'm withholding in the current draft (marked as \hl{references} in color). 
  \item The general discussion should be expanded and polished. In particular concerning the discussion of lexical representations and the language of thought approach;
  \item I still want to add references and clarify the discussion on homogeneous populations and frequency effects in the discussion of the model's predictions;
  \item The analysis may benefit from further data and less of a qualitative overview of the parameters' effect (e.g. the regression analysis Michael conducted);
  \item As you will have noticed when reading footnote $3$ I think the performance of $L_6$ being on par with $L_4$ is somewhat problematic;
  \item There's a bit of repetition due to the early stages of the draft and the fact that scalar implicatures and some matrices are discussed in non-contiguous parts of the paper. Whether they stay depends on whether we roughly keep the current structure and length;
  \item In principle, one could make a direct comparison between the case-study of \citet{kalish+griffiths:2007} on compositionality using iterated learning and iterated learning + functional pressure. This could be relegated to the appendix but I think this is somewhat tangential and wouldn't add much. The same goes for an explicit comparison of {\em some} vs. {\em some but not all} using a language of thought-style representation. I currently decided against it, but let me know if you see worth in either option.
\end{itemize}

>>>>

\subsection{Discussion}
Broadly speaking these results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population provided there is a bias for simpler representations. This prediction hinges on three assumptions. First, that language is pressured toward both expressivity and learnability. Second, that language use is not too unpredictable -- low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn. Third, that learners prefer simpler over more complex lexical representations. An important addendum to this third condition being that a combination of rationality in choice and maximization in learning requires a weaker bias towards simplicity. Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.  This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer.

While a major share of non upper-bounded lexical meanings is predicted by the literature, it is less clear to what extent other types should be present in the final population, if at all. It seems reasonable to expect functionally suboptimal types $L_1$, $L_2$ and $L_3$ to be ruled out because they fail to enable their users to communicate effectively. However, this is not true of $L_4$.\footnote{$L_6$ presents a special case. In our current setup, it mirrors $L_5$ in enabling for the pragmatic narrowing of message $m_2$ rather than $m_1$. However, this association of $s_2$ with $m_2$ under favorable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$. $L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'. Our representation of lexica as Boolean matrices is blind to this anomaly.} Notwithstanding, the prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is is not implausible (\hl{references on individual level differences in the derivation of scalar implicatures}). Alternatively, a stronger tendency for posterior maximization has to be assumed (cf. Figure \ref{fig:prior-posterior}). This empirical issue relates to other two aspects left undiscussed in our present analysis: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences. We tacitly assumed pragmatic reasoning to come at no cost. However, there is experimental evidence that suggests  that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf. \citealt{deNeys+schaeken:2007, huang+snedeker:2009}). This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations. Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used. We conject that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line. This opens a possible venue to address our previous question about the expected presence of $L_4$-like semantics.

\section{General discussion}
We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of learning biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}. This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on language. We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in learnability: Simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred. This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences fail to lexicalize.

The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning. The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}. In contrast to our proposal, however, Kirby et al. model expressivity as exerting its force only in the production of learning data. This model's expressivity parameter thereby fulfills a similar role  to high values of $\lambda$ and $\alpha$ in making speaker behavior more deterministic. In this way, it ``favors'' unambiguous languages. Crucially, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration. Our reasoning behind its inclusion lies in the empirical and theoretical observation that learnability alone can have the undesirable outcome of favoring functionally defective languages. This outcome is predicted by our analysis and similar effects have also been reported in a number laboratory experiments where the participants' task was to learn and subsequently reproduce the language produced by a previous participant. This process was shown to generally lead to a proliferation of languages that associated a large number of meanings with a single form (\hl{references}). In contrast, adding an experiments involving an interactive component have been found to foster languages that enable interlocutors to distinguish meanings accurately  (\hl{references}; for a review of these studies and further references see \citealt{kirby+etal:2015, tamariz+kirby:2016}). More generally, we think there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer. The former solely depends on the mechanism by which speakers associate form and meaning. The latter additionally hinges on the hearers linguistic experience and her capability to interpret linguistic input based on this experience. In sum, we contend that successful information transfer in a linguistic community is central the adoption of a communication system and that this measure is not faithfully reflected by taking only production into consideration.

One of our central  assumptions is the learning bias for simpler lexical representations. Albeit implicit in our formalization, the conceptual foundation for a compression preference at the lexical level is grounded in work in the language of thought tradition put forward by \citet{piantadosi+etal:2012a,piantadosi+etal:underreview}. Following Piantadosi and colleagues we assumed that lexical meanings are expressed in a representational language and that the learners' task is to infer lexica that best explain the data witnessed. Furthermore, we assumed learners to not only infer lexical meaning but types of linguistic use. Whether language use is culturally transmitted is an empirical matter that can not be settled here. However, this idealized model suggests that semantic structure evinced cross-linguistically that lends itself for pragmatic exploitation reflects lexical information that is both easy to learn and, perhaps more interestingly, also easy to learn how to exploit via mutual reasoning.

\section{Conclusion}
Language change is affected by intertwined pressures. We investigated the selection of lexical meanings lacking upper-bounds in a case study on role of semantics and pragmatics in satisfying such pressures. We showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use. That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages fail to lexicalize certain pragmatic inferences. More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interaction as well as their representational complexity and learnability.

@@

\section{Some thoughts + what may be missing}
\begin{itemize}
  \item I still need to review, read and find some references I'm withholding in the current draft (marked as ``\hl{references}'' in color). 
  \item The discussion (sub) sections should be expanded and polished. In particular the discussion concerning lexical representations and the language of thought. I'm aware that this part is not as readable as the rest;
  \item The analysis may benefit from further data and less of a qualitative overview of the parameters' effect (e.g. the regression analysis Michael conducted);
  \item As you will have noticed when reading footnote $3$, I think the performance of $L_6$ being on par with $L_4$ is somewhat problematic;
  \item There's a fair share of repetition, this is partly due to the fact that scalar implicatures and some matrices are discussed in non-contiguous parts of the paper. Whether the repetition is necessary depends on whether we roughly keep the current structure and length or change things;
  \item In principle, we could make a direct comparison between the case-study of \citet{griffiths+kalish:2007} on compositionality using iterated learning and iterated learning + functional pressure. This could be relegated to the appendix. However, I'm inclined to think this is somewhat tangential and wouldn't add much. The same goes for an explicit comparison of {\em some} vs. {\em some but not all} using a language of thought-style representation. Let me know if you see worth in either option.
\end{itemize}
